---
title: "Gapminder Report"
author: "Tammie Tam"
date: "7/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# load packages
library(tidyverse)
library(corrr)
library(plotly)
```

1. Read in the `gapminder_clean.csv` data as a `tibble` using `read_csv`.
```{r}
# load in gapminder dataset
gapminder <- read_csv("gapminder_clean.csv")
head(gapminder)
```

2. Filter the data to include only rows where `Year` is `1962` and then make a scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap` for the filtered data.
```{r}
# filter only Year 1962 
gapminder_1962 <- gapminder %>% filter(Year == "1962")

# plot gdp v CO2
gapminder_1962 %>% ggplot(aes(gdpPercap,`CO2 emissions (metric tons per capita)`)) + geom_point()
```

3. On the filtered data, calculate the correlation of `'CO2 emissions (metric tons per capita)'` and `gdpPercap`. What is the correlation and associated p value?
```{r}
gdp.co2.cor <- cor.test(x=gapminder_1962$gdpPercap,y=gapminder_1962$`CO2 emissions (metric tons per capita)`)
# correlation val
gdp.co2.cor$estimate
# correlation p val
gdp.co2.cor$p.value
```

4. On the unfiltered data, answer "In what year is the correlation between `'CO2 emissions (metric tons per capita)'` and `gdpPercap` the strongest?" Filter the dataset to that year for the next step...
```{r}
gapminder %>% group_by(Year) %>% summarize(correlation=cor(`CO2 emissions (metric tons per capita)`,gdpPercap,use = "na.or.complete")) %>% subset(correlation == max(correlation))
gapminder_1967 <- gapminder %>% filter(Year=="1967")
```

5. Using `plotly`, create an interactive scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap`, where the point size is determined by `pop` (population) and the color is determined by the `continent`. You can easily convert any `ggplot` plot to a `plotly` plot using the `ggplotly()` command.
```{r}
plt <- gapminder_1967 %>% ggplot(aes(gdpPercap,`CO2 emissions (metric tons per capita)`, color = continent, size = pop)) + geom_point()
ggplotly(plt)
```

Now, without further guidance, use your `R` Data Science skills (and appropriate statistical tests) to answer the following:

1. What is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`? (stats test needed)
2. Is there a significant difference between Europe and Asia with respect to `'Imports of goods and services (% of GDP)'` in the years after 1990? (stats test needed)
3. What is the country (or countries) that has the highest `'Population density (people per sq. km of land area)'` across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)
4. What country (or countries) has shown the greatest increase in `'Life expectancy at birth, total (years)'` since 1962? 

