---
title: "Gapminder Report"
author: "Tammie Tam"
date: "7/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# load packages
library(tidyverse)
library(corrr)
library(plotly)
```

1. Read in the `gapminder_clean.csv` data as a `tibble` using `read_csv`.
```{r, message=FALSE}
# load in gapminder dataset
gapminder <- read_csv("gapminder_clean.csv") %>% column_to_rownames(., var = 'X1')
head(gapminder)
```

2. Filter the data to include only rows where `Year` is `1962` and then make a scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap` for the filtered data.
```{r, message=FALSE}
# filter only Year 1962 
gapminder_1962 <- gapminder %>% filter(Year == "1962")

# plot gdp v CO2
gapminder_1962 %>% ggplot(aes(gdpPercap,`CO2 emissions (metric tons per capita)`)) + geom_point() +  labs(title = "CO2 emissions by gdpPercap,continent, population size")
```

3. On the filtered data, calculate the correlation of `'CO2 emissions (metric tons per capita)'` and `gdpPercap`. What is the correlation and associated p value?
```{r}
gdp.co2.cor <- cor.test(x=gapminder_1962$gdpPercap,y=gapminder_1962$`CO2 emissions (metric tons per capita)`)
# correlation val
gdp.co2.cor$estimate
# correlation p val
gdp.co2.cor$p.value
```

4. On the unfiltered data, answer "In what year is the correlation between `'CO2 emissions (metric tons per capita)'` and `gdpPercap` the strongest?" Filter the dataset to that year for the next step...
```{r}
gapminder %>% group_by(Year) %>% summarize(correlation=cor(`CO2 emissions (metric tons per capita)`,gdpPercap,use = "na.or.complete")) %>% subset(correlation == max(correlation))
gapminder_1967 <- gapminder %>% filter(Year=="1967")
```

5. Using `plotly`, create an interactive scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap`, where the point size is determined by `pop` (population) and the color is determined by the `continent`. You can easily convert any `ggplot` plot to a `plotly` plot using the `ggplotly()` command.
```{r}
plt <- gapminder_1967 %>% ggplot(aes(gdpPercap,`CO2 emissions (metric tons per capita)`, color = continent, size = pop)) + geom_point() + labs(title = "CO2 emissions by gdpPercap,continent, population size", color = "continent", size = "")
ggplotly(plt)
```

Now, without further guidance, use your `R` Data Science skills (and appropriate statistical tests) to answer the following:

1. What is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`? (stats test needed)

The Energy usage between continent is significantly different (p = 2e-16 < 0.05).
```{r}
gapminder_contE <- gapminder %>% select(continent, `Energy use (kg of oil equivalent per capita)`) %>% drop_na()

contE_aov <- aov(`Energy use (kg of oil equivalent per capita)` ~ continent, data = gapminder_contE)
summary(contE_aov)
```

2. Is there a significant difference between Europe and Asia with respect to `'Imports of goods and services (% of GDP)'` in the years after 1990? (stats test needed)

There is no significant difference (p = 0.1776 > 0.05) between Europe and Asia's Import of goods and services (% of GDP) after 1990.
```{r}
gapminder_1990 <- gapminder %>% filter(Year > "1990" & continent %in% c("Asia", "Europe")) %>% select(continent, `Imports of goods and services (% of GDP)`) %>% drop_na()

t.test(`Imports of goods and services (% of GDP)` ~ continent, data = gapminder_1990)
```

3. What is the country (or countries) that has the highest `'Population density (people per sq. km of land area)'` across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)
```{r}
gapminder %>% group_by(`Country Name`) %>% summarize(mean_popdensity = mean(`Population density (people per sq. km of land area)`)) %>% drop_na() %>% subset(mean_popdensity == max(mean_popdensity))
```

4. What country (or countries) has shown the greatest increase in `'Life expectancy at birth, total (years)'` since 1962? 
```{r}
gapminder %>% select(`Country Name`, Year, `Life expectancy at birth, total (years)`) %>% group_by(`Country Name`) %>% pivot_wider(names_from = Year, values_from = `Life expectancy at birth, total (years)`) %>% select(`Country Name`, `1962`, `2007`) %>% mutate(change_lifeexp = `2007` - `1962`) %>% drop_na() %>% subset(change_lifeexp == max(change_lifeexp))
```
